An Experimental Evaluation of Hybrid Query Algorithms

### 2025-04-01 VLDB2026/main.tex
Create main.tex

%%
%% The "author" command and its associated commands are used to define the authors and their affiliations.
% \author{Ben Trovato}
% \affiliation{%
%   \institution{Institute for Clarity in Documentation}
%   \streetaddress{P.O. Box 1212}
%   \city{Dublin}
%   \state{Ireland}
%   \postcode{43017-6221}
% }
% \email{trovato@corporation.com}

% \author{Lars Th{\o}rv{\"a}ld}
% \orcid{0000-0002-1825-0097}
% \affiliation{%
%   \institution{The Th{\o}rv{\"a}ld Group}
%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%   \city{Hekla}
%   \country{Iceland}
% }
% \email{larst@affiliation.org}

% \author{Valerie B\'eranger}
% \orcid{0000-0001-5109-3700}
% \affiliation{%
%   \institution{Inria Paris-Rocquencourt}
%   \city{Rocquencourt}
%   \country{France}
% }
% \email{vb@rocquencourt.com}

% \author{J\"org von \"Arbach}
% \affiliation{%
%   \institution{University of T\"ubingen}
%   \city{T\"ubingen}
%   \country{Germany}
% }
% \email{jaerbach@uni-tuebingen.edu}
% \email{myprivate@email.com}
% \email{second@affiliation.mail}

% \author{Wang Xiu Ying}
% \author{Zhe Zuo}
% \affiliation{%
%   \institution{East China Normal University}
%   \city{Shanghai}
%   \country{China}
% }
% \email{firstname.lastname@ecnu.edu.cn}

% \author{Donald Fauntleroy Duck}
% \affiliation{%
%   \institution{Scientific Writing Academy}
%   \city{Duckburg}
%   \country{Calisota}
% }
% \affiliation{%
%   \institution{Donald's Second Affiliation}
%   \city{City}
%   \country{country}
% }
% \email{donald@swa.edu}
\author{Ben Trovato$^1$, Lars Th{\o}rv{\"a}ld$^2$, Valerie B\'eranger$^3$, J\"org von \"Arbach$^3$, Wang Xiu Ying$^3$, Donald Fauntleroy Duck$^3$}

\affiliation{%
  \institution{1. Institute for Clarity in Documentation, Ireland}
  \institution{2. The Th{\o}rv{\"a}ld Group, Iceland}
  \institution{3. Inria Paris-Rocquencourt, France}
}

\email{trovato@corporation.com, larst@affiliation.org, vb@rocquencourt.com, jaerbach@uni-tuebingen.edu, firstname.lastname@ecnu.edu.cn, donald@swa.edu}

@@

Recent studies have shown that hybrid queries, including attribute filtering and range filtering, hold significant practical value. However, current evaluation efforts lack a unified benchmark and comprehensive testing. Existing research has failed to cover mainstream algorithms and has not systematically compared or deeply analyzed different methods. To address this issue, we have designed a complete evaluation framework for hybrid queries. Our study introduces more than ten hybrid query algorithms and systematically classifies them based on multiple dimensions such as index structure and filtering strategy, providing a reference for the categorization of hybrid queries. In the experimental section, for attribute filtering, we constructed standardized attribute sets, enabling a unified comparison of algorithms in terms of index construction efficiency, query performance, and robustness. For range filtering, we systematically tested the algorithms' performance in these three aspects by varying the query ranges. Additionally, we conducted an in-depth analysis of the experimental results based on the underlying principles of the algorithms. The extensive experimental results reveal the strengths and weaknesses of each algorithm. Based on these findings, we have developed a set of practical guidelines for algorithm selection, offering reliable references for different application scenarios. Furthermore, we have identified potential directions for improvement to address the current limitations of these algorithms.
\vspace{-0.2cm}
\end{abstract}

@@

Nearest Neighbor (NN) search \cite{cover1967nearest} aims to find the closest vector in a given space and serves as a fundamental algorithm in vector retrieval, with widespread applications in recommendation systems  \cite{monolith, Airbnb, Related_Pins_at_Pinterest, wang2018billion, pal2020pinnersage, okura2017embedding}  and image retrieval \cite{wang2012scalable}.
% 需要添加引用，下面这段
However, with the exponential growth of data volume and the increasing dimensionality of vectors \cite{weber1998quantitative}, traditional NN search methods struggle to meet the demands of real-time search. To address this issue, researchers have turned to more efficient approaches—Approximate Nearest Neighbor search ~\cite{arya1998optimal,beis1997shape,gionis1999similarity,malkov2018efficient}. ANN algorithms significantly improve search efficiency by constructing effective indexing structures, albeit at the cost of reduced accuracy.

Nevertheless, as application scenarios grow increasingly complex, simple ANN search can no longer satisfy all practical needs. For instance, as shown in Figure\ref{fig:hybrid ANNS}, users on e-commerce platforms may search for clothing items by retrieving visually similar products based on an image. Additionally, users may impose further requirements such as price, color, or brand preferences. Such scenarios necessitate a retrieval system capable of simultaneously addressing vector similarity (e.g., product image) and attribute constraints (e.g., brand name) \cite{tian2023approximate}. When the constraint involves a specific attribute value, this problem is referred to as Attribute Filtering Approximate Nearest Neighbor (AF-ANN) search ~\cite{NHQ,Filtered-diskann} . For example, a user may seek a green piece of clothing. If the constraint involves a range condition, it is termed Range Filtering Approximate Nearest Neighbor (RF-ANN) search  ~\cite{serf,iRangeGraph}. For instance, filtering clothes priced between $100 and $200. To meet these application demands, hybrid query \cite{JD-e-commerce, analyticdb} techniques have emerged. Hybrid queries integrate vector retrieval with conditional filtering, optimizing their interaction to significantly enhance efficiency and flexibility in complex query scenarios.

% However, as application scenarios become more complex, simple ANN search is no longer sufficient to meet all practical needs. New requirements have emerged for ANN search systems. For example, as Figure \ref{fig:hybrid ANNS}, a user on an e-commerce platform may search for a dress by retrieving similar items based on an image. Additionally, the user can apply structured attribute filters such as price, color, or brand. This scenario necessitates a retrieval system that satisfies both vector similarity and attribute constraints \cite{tian2023approximate}. When the constraint is a Boolean condition—e.g., the user specifically requires a green dress—the problem is referred to as Attribute Filtering Approximate Nearest Neighbor (AF-ANN) search ~\cite{NHQ,Filtered-diskann} problem. If the constraint is a range condition, such as filtering dresses priced between 100 and 200, it becomes the Range Filtering Approximate Nearest Neighbor (RF-ANN) search ~\cite{serf,iRangeGraph} problem. To address such cases, hybrid query \cite{JD-e-commerce, analyticdb} has been proposed. It jointly optimizes vector retrieval and conditional filtering. This approach significantly improves efficiency and flexibility in complex query scenarios.
\begin{figure*}
    \centering
    \setlength{\abovecaptionskip}{0cm}
    \setlength{\belowcaptionskip}{-0.4cm}
    \includegraphics[width=\textwidth]{figures/hybrid ANNS.pdf}
    \caption{Hybrid query example}
    % \vspace{-10pt} % 调整这个值来缩小空隙
    \label{fig:hybrid ANNS}
\end{figure*}
\subsection{Motivation}
% 近年来，混合查询算法发展迅速，已有多种属性过滤或范围过滤工作被提出。尽管 BigANN 2023 大赛\cite{bigann2023}的 filter 赛道对部分算法进行了性能测试，但其评估仍存在局限性：1) 参赛算法数量有限，未能覆盖当前主流方法；2) 仅评估单属性和双属性查询，未涉及更复杂的多属性场景；3) 测试数据集的属性分布不够全面。此外，不同算法在索引构建的时空开销、查询性能随属性数量变化的表现、以及对不同属性分布的适应性等方面尚缺乏系统性评估。目前，大多数研究仅关注属性过滤或范围查询的单一问题，而未能全面衡量混合查询的整体性能。为此，我们基于混合查询进行完整实验评估，全面分析不同算法在索引构建成本、查询效率及鲁棒性等方面的表现。
In recent years, hybrid query algorithms have developed rapidly, giving rise to numerous attribute filtering ~\cite{NHQ,diskann} and range-filtering algorithms \cite{serf,iRangeGraph}. In practical applications, the performance of hybrid query algorithms is influenced not only by unstructured data but also closely related to structured data. For attribute filtering algorithms, factors such as the number of base attributes, the number of query attributes, attribute distribution \cite{UNG}, and attribute selectivity \cite{ACORN} significantly impact algorithm performance. As for range filtering algorithms, the size of the query range and the characteristics of different datasets also play an important role in determining algorithm performance.

% In recent years, hybrid query algorithms have advanced rapidly, with numerous algorithms  proposed for attribute filtering and range filtering. 
% In practice, the performance of hybrid query algorithms is also influenced by the characteristics of structured data. For attribute filtering algorithms, factors such as the number of base attributes, the number of query attributes, attribute distributions \cite{UNG}, and attribute selectivity \cite{ACORN} all affect actual performance. For range filtering algorithms, query range size and different datasets will affect the performance of the algorithm.
Despite BigANN 2023 \cite{bigann2023}  evaluating the performance of some algorithms, its study still has several notable limitations:
1)The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. For instance, NHQ and Filtered-DiskANN in attribute filtering algorithms have not been assessed.
2)It focuses only on single-attribute and dual-attribute query scenarios, neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes. For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range, etc.) simultaneously when searching.
3)It does not evaluate algorithm performance under different attribute distributions. However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.

Moreover, aside from BigANN, there is currently a lack of systematic evaluation specifically targeting hybrid query algorithms. To fill this research gap, we conducted a comprehensive experimental evaluation of hybrid query algorithms, analyzing the index construction costs and query efficiency across different scenarios. Additionally, we further assessed the robustness of each algorithm under varying conditions.

% Although the filter track of the BigANN 2023 competition \cite{bigann2023} evaluated the performance of some algorithms, its assessment has several limitations: 
% 1) The number of evaluated algorithms is limited, failing to cover mainstream methods. Mainstream attribute filtering algorithms such as NHQ and Filtered-DiskANN were not evaluated.
% 2) The evaluation focuses only on single-attribute and dual-attribute queries, without considering more complex multi-attribute scenarios. However, in actual applications, an object often involves multiple attributes, and the same is true for queries. For example, in e-commerce scenarios, users often specify multiple attributes (such as color, brand, price range, etc.) for search at the same time.
% 3) It did not include algorithm performance under different attribute distributions. Different distribution of attributes may have a direct impact on the filtering strategy and index construction of the hybrid query algorithm.
% In addition, there is no relevant research to summarize and evaluate the relevant algorithms for hybrid queries.

% Furthermore, with the exception of the BigANN competition, there has been no dedicated study that provides a comprehensive review and evaluation of hybrid query algorithms. 
% In addition, apart from the BigANN competition, no dedicated evaluation of hybrid query algorithms exists. To fill this gap, we conduct a comprehensive experimental evaluation of hybrid query algorithms, analyzing index construction overhead and query efficiency across diverse scenarios. In addition, we assess the robustness of each algorithm.

% Moreover, there has been no systematic evaluation of different algorithms in terms of index construction overhead, query performance across varying attribute numbers, and adaptability to different attribute distributions. Most existing studies focus solely on attribute filtering or range filtering, without providing a holistic assessment of hybrid queries. To address this gap, we conduct a comprehensive experimental evaluation of hybrid queries, analyzing the indexing overhead, query efficiency, and robustness of different algorithms.
% In recent years, hybrid query algorithms have advanced rapidly, but their differences in attribute distribution adaptability and multi-attribute query performance have not been systematically evaluated. Although the filter track of the BigANN 2023 competition\cite{bigann2023} tested the performance of some algorithms, its evaluation scope has several limitations. First, the number of participating algorithms was limited, failing to cover the mainstream hybrid query algorithms. Second, the test dataset lacked diversity in attribute cardinality and attribute distribution, which may not fully reflect algorithm performance under different data characteristics. Third, the evaluation focused only on single-attribute and two-attribute query, without considering more complex multi-attribute scenarios, limiting the generalizability and practical value of the results. Moreover, the competition did not include evaluations of range filtered algorithms.

% Based on this background, we aim to design a series of systematic experiments. We will construct diverse datasets for attribute filtered query and comprehensively evaluate mainstream hybrid query algorithms.
\subsection{Our Contributions}
% The core contribution of this study lies in the systematic evaluation of hybrid query algorithms, which specifically includes the following aspects:
% We focus on ANN search in hybrid query scenarios. We systematically review and evaluate existing algorithms and systems. Our main contributions are summarized in four parts:
This study focuses on the problem of ANN search in hybrid query scenarios and provides a comprehensive review and evaluation of existing algorithms and systems. The main contributions are summarized in the following four aspects.

(1)\textbf{ Systematic Classification and Overview.}
We systematically classify more than ten representative attribute filtering algorithms from multiple dimensions, including index structure, filtering strategy, Boolean logic support, and index construction methods. Additionally, we survey range filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts provide a unified reference framework for future research.

(2)\textbf{ Evaluation of Attribute-Filtering Algorithms.}
% We select several representative algorithms for Attribute Filtered ANN Search. Using six real-world datasets, we evaluate their performance under different settings, such as the number of attributes, attribute distribution, and selectivity. We analyze their performance in terms of index construction time, index size,peak memory usage, query speed (QPS), and query accuracy. We also discuss their robustness and adaptability in complex scenarios.
We experimentally evaluate 13 attribute filtering ANN algorithms on six real-world datasets. By analyzing the performance variations of these algorithms under different numbers of attributes, attribute distributions, and selectivity conditions, we reveal the strengths and weaknesses of each algorithm. Furthermore, we explore the robustness and adaptability of these algorithms under complex query conditions. Evaluation metrics include index construction time, index size, peak memory usage,QPS and search accuracy.

(3)\textbf{ Evaluation of Range-Filtering Algorithms.}
% We evaluate five main algorithms (SeRF, WinFilter, iRangeGraph, DSG, and UNIFY) for Range Filtered ANN Search. We run experiments on three large-scale datasets with different query range settings. We compare their performance in index construction, storage usage, and query efficiency. We also study the effect of index structures on algorithm performance. This provides useful insights for future improvements.
We benchmark five mainstream range filtering ANN algorithms on three large-scale datasets, using varying query range settings in the experiments. The results demonstrate the performance of these algorithms in terms of index construction efficiency, storage overhead, and query performance. Additionally, we conduct an in-depth analysis of how index structures impact algorithm performance, providing valuable insights for algorithm design.

(4)\textbf{ Recommendations and Challenges.}
Based on the experimental results, we provide algorithm selection recommendations for common application scenarios and highlight key challenges in the field of hybrid queries. These challenges include limited Boolean logic support, the lack of multi-attribute range filtering capabilities, the high indexing costs of graph-based methods, and the sensitivity of algorithms to data distribution. Currently, few methods simultaneously support both attribute filtering and range filtering, pointing to potential directions for future research.

@@

However, to address the curse of dimensionality faced by NN search \cite{dimcurse}, existing researches focus on approximate solutions, known as ANN  search. We typically use $\text{Recall}@k = \frac{|R \cap \hat{R}|}{k}$ to evaluate the accuracy of ANN search algorithms, where $R$ denotes the true top-$k$ nearest neighbors of the query, $\hat{R}$ denotes the approximate top-$k$ nearest neighbors returned by the ANN search algorithm.

As the complexity of real-world application requirements increases, NN search has evolved into hybrid NN search with attribute constraints. Depending on the nature of the attribute constraints, hybrid NN search can be divided into two categories: 1) Attribute Filtering Nearest Neighbor (\textit{AF-NN}) search. 2) Range Filtering Nearest Neighbor (\textit{RF-NN}) search. We provide their formal definitions below.

@@

\subsection{Index Structures in ANN Search}

Current hybrid query methods mainly adopt graph-based ~\cite{nsw,kgraph,nsg,fanng,ngt} or Inverted File Index (IVF)-based \cite{PQ} index structures. We briefly introduce the fundamental principles of these two types in the following.

\vspace{1em}
\noindent\textbf{\underline{Graph.}}
Graph-based ANN search algorithms accelerate queries by building graph indexes on datasets. As illustrated in Figure~\ref{fig:graph}, each data point in the dataset corresponds to a point in the graph, and neighboring vertices (e.g., $a$ and $b$) are connected via edges based on their distance $\textit{dist}(a, b)$. In the graph index, each point maintains connections only to its nearest neighbors to preserve query efficiency. For instance, in Figure~\ref{fig:graph}, the gray point $c$ is connected to five neighbors $(a, b, d, e, f)$ and can traverse to them via corresponding edges.

Given a query point $q$, the graph-based ANN search typically follows a greedy search strategy to find the $k$ nearest neighbors. For illustration, consider the case where $k = 1$ (i.e., finding the single nearest neighbor of $q$): The algorithm initiates the search from an entry point (the gray point $c$). If there exists a neighbor $n$ (the point $f$) of the current point such that $\textit{dist}(n, q) < \textit{dist}(c, q)$, the algorithm updates the current point to $n$. This process is iteratively repeated—always moving to the neighbor closest to $q$—until no neighbor is found that is closer than the current point. The final point reached (the blue point $j$) is returned as the approximate nearest neighbor of $q$.

@@

We provide an overview and analysis of over ten hybrid query methods, all drawn from recent research efforts, and collectively referred to as algorithms in the following text. Specifically, Section~3.1 describes methods based on attribute filtering, Section~3.2 discusses those based on range filtering, and Section~3.3 introduces vector libraries and databases that support both attribute and range filters.

\vspace{-10pt}
\subsection{Attribute Filtering Algorithms}

We summarize six attribute filtering algorithms and provide an intuitive comparison in Table~\ref{tab:compair_1}.

\vspace{1em}
\noindent\textbf{\underline{NHQ \cite{NHQ}.}}  
Traditional attribute filtering algorithms usually perform attribute constraints and ANN search separately. In contrast, NHQ is the first to implement simultaneous filtering, integrating both aspects into a unified framework. NHQ constructs an index based on a nearest neighbor graph and introduces a fusion distance that jointly captures vector similarity and attribute similarity. Leveraging this fusion distance, NHQ unifies vector similarity and attribute matching into a single comprehensive similarity measure and builds the graph accordingly. During the query process, NHQ efficiently prunes irrelevant edges via this composite index, enabling fast retrieval of results that satisfy both vector similarity and attribute constraints.

@@

Filtered-DiskANN also supports simultaneous filtering. Built upon the Vamana \cite{diskann} graph-based ANN index, it incorporates attribute information directly into the graph during index construction. This integration ensures that the index reflects both vector proximity and attribute constraints. Moreover, Filtered-DiskANN supports SSD-based storage, improving scalability.

Filtered-DiskANN proposes two index variants:  
1) \textbf{FilteredVamana}, which incrementally builds the graph index by inserting data points and dynamically adding edges, allowing adaptive expansion.  
2) \textbf{StitchedVamana}, which adopts a batch construction strategy—constructing separate Vamana subgraphs for each attribute, followed by merging and edge pruning.

@@

ACORN is built upon HNSW \cite{hnsw}, a hierarchical graph-based ANN index. It supports high-cardinality and unrestricted predicate sets, overcoming the limitations of methods restricted to small-scale equality filters. It constructs a denser hierarchical graph by expanding node neighborhoods and prunes edges at lower levels to control index size. During querying, ACORN eliminates unnecessary distance computations by filtering out nodes that violate attribute constraints. This effectively maintains a nearest neighbor graph over valid nodes only. 
%During the query process, ACORN avoids unnecessary distance computations by filtering out nodes that do not meet attribute constraints, effectively simulating a nearest neighbor graph over valid nodes only.

ACORN includes two index variants:  
1) \textbf{ACORN-$\gamma$}, which expands neighbor lists during construction, increasing memory usage for higher performance.  
2) \textbf{ACORN-1}, which extends neighbor lists by considering second-hop neighbors during search, reducing index size with minor performance loss.

\vspace{1em}
\noindent\textbf{\underline{UNG \cite{UNG}.}} 
Graph-based methods(NHQ, Filtered-DiskANN, ACO-\\RN)do not guarantee result completeness, while CAPS lacks robustness when base and query vectors differ in attribute cardinality. UNG is proposed to overcome both limitations.

@@

UNG supports Boolean filtering with two modes:  
1) The query attribute set is a subset of the data attribute set.  
2) The query attribute set exactly matches the data attribute set.

@@

\subsection{Range Filtering Algorithms}

We introduce five representative range filtering algorithms and present a comparative overview in Table~\ref{tab:range_algo}.

>>>>

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}

\author{xxxxxxxx}
\affiliation{%
 \institution{Huazhong University of Science and Technology}
}
%%
%% The "author" command and its associated commands are used to define the authors and their affiliations.

% \author{Mengzhao Wang$^1$, Xiaoliang Xu$^1$, Qiang Yue$^1$, Yuxiang Wang$^{1,*}$}
% \affiliation{%
%  \institution{$^1$Hangzhou Dianzi University, China}
% }
% \email{{mzwang,xxl,yq,lsswyx}@hdu.edu.cn}
% \author{Ben Trovato$^1$, Lars Th{\o}rv{\"a}ld$^2$, Valerie B\'eranger$^3$, J\"org von \"Arbach$^3$, Wang Xiu Ying$^3$, Donald Fauntleroy Duck$^3$}

% \affiliation{%
%   \institution{1. Institute for Clarity in Documentation, Ireland}
%   \institution{2. The Th{\o}rv{\"a}ld Group, Iceland}
%   \institution{3. Inria Paris-Rocquencourt, France}
% }

% \email{trovato@corporation.com, larst@affiliation.org, vb@rocquencourt.com, jaerbach@uni-tuebingen.edu, firstname.lastname@ecnu.edu.cn, donald@swa.edu}

@@

Recent studies demonstrate the significant practical value of hybrid queries, which integrate vector search with structured filters (e.g., attribute and range filtering) for refined retrieval.
However, current evaluations lack unified benchmarking standards and systematic assessment methodologies. Existing studies not only fail to cover mainstream algorithms but also omit systematic comparisons or in-depth analyses of different methods. To address this issue, we design a complete evaluation framework for hybrid queries. Our study introduces 15 hybrid query algorithms and systematically classifies them based on multiple dimensions such as index organization and filtering strategy, providing a reference for the categorization of hybrid queries. In the experiments, for attribute filtering, we construct standarded attribute sets, enabling a unified comparison of algorithms in terms of index construction efficiency, query performance, and robustness. For range filtering, we systematically evaluate algorithm performance across these 3 metrics through controlled variation of query ranges. Additionally, we conduct an in-depth analysis of the experimental results based on the underlying principles of the algorithms. The extensive experimental results reveal the strengths and weaknesses of each algorithm. Based on these findings, we develop a set of practical guidelines for algorithm selection, offering reliable references for different application scenarios. Furthermore, we identify potential directions for improvement to address the current limitations of these algorithms.
\vspace{-0.2cm}
\end{abstract}

@@

Nearest Neighbor (NN) search \cite{cover1967nearest} aims to find the closest vector in a given space and serves as a fundamental algorithm in vector retrieval, with widespread applications in recommendation systems  \cite{monolith, Airbnb, Related_Pins_at_Pinterest, wang2018billion, pal2020pinnersage, okura2017embedding}  and image retrieval \cite{wang2012scalable}.
% 需要添加引用，下面这段
However, with the exponential growth of data volume and the increasing dimensionality of vectors \cite{weber1998quantitative}, traditional NN search methods struggle to meet the demands of real-time search. To address this issue, researchers turn to more efficient approaches - Approximate NN search ~\cite{arya1998optimal,beis1997shape,gionis1999similarity,malkov2018efficient}. ANN algorithms significantly improve search efficiency by constructing effective indexing structures, albeit at the cost of reduced accuracy.

Nevertheless, as application scenarios grow increasingly complex, simple ANN search can no longer satisfy all practical needs. For instance, Figure ~\ref{fig:hybrid ANNS} illustrates a case where users on e-commerce platforms search for clothing items by retrieving visually similar products based on an image. Additionally, users may impose further requirements such as price, color, or brand preferences. Such scenarios necessitate a retrieval system capable of simultaneously addressing vector similarity (e.g., product image) and attribute constraints (e.g., brand name) \cite{tian2023approximate}. When the constraint involves a specific attribute value, this problem refers to Attribute Filtering Approximate Nearest Neighbor (AF-ANN) search ~\cite{NHQ,Filtered-diskann} . For example, a user may seek a green piece of clothing. If the constraint involves a range condition, the problem is known as Range Filtering Approximate Nearest Neighbor (RF-ANN) search ~\cite{serf,iRangeGraph}. For instance, filtering clothes priced between 100 and 200. To meet these application demands, hybrid query \cite{JD-e-commerce, analyticdb} techniques emerge. Hybrid queries integrate vector retrieval with conditional filtering, optimizing their interaction to significantly enhance efficiency and flexibility in complex query scenarios.

\begin{figure*}
    \centering
    \setlength{\abovecaptionskip}{0cm}
    \setlength{\belowcaptionskip}{-0.4cm}
    \includegraphics[width=\textwidth]{figures/hybrid ANNS.pdf}
    \caption{Hybrid query example}
    % \vspace{-10pt} % 调整这个值来缩小空隙
    \label{fig:hybrid ANNS}
\end{figure*}
\subsection{Motivation}

In recent years, hybrid query algorithms develop rapidly, giving rise to numerous attribute-filtering ~\cite{NHQ,diskann} and range-filtering algorithms \cite{serf,iRangeGraph}. In practical applications, the performance of hybrid query algorithms is influenced not only by unstructured data but also closely related to structured data. For attribute-filtering algorithms, factors such as the number of base attributes, the number of query attributes, attribute distribution \cite{UNG}, and attribute selectivity \cite{ACORN} significantly impact algorithm performance. As for range-filtering algorithms, the size of the query range and the characteristics of different datasets also play an important role in determining algorithm performance.

Despite BigANN 2023 \cite{bigann2023} evaluates the performance of some algorithms, it presents several notable limitations:
1) The evaluation covers a limited number of algorithms and fails to comprehensively include mainstream methods. For instance, attribute-filtering algorithms such as NHQ and UNG are not included in the evaluation. 2) It focuses only on single-attribute and dual-attribute query scenarios, neglecting the more complex demands of multi-attribute queries. In real-world applications, an object typically involves multiple attributes. For example, in e-commerce scenarios, users often specify multiple attributes (e.g., color, brand, price range) simultaneously when searching.
3) It does not evaluate algorithm performance under different attribute distributions. However, variations in attribute distribution may directly affect the filtering strategies and index construction efficiency of hybrid query algorithms.

Moreover, aside from BigANN, there is currently a lack of systematic evaluation specifically targeting hybrid query algorithms. To fill this gap, we conduct a comprehensive experimental evaluation of hybrid query algorithms, analyzing the index construction costs and query efficiency across different scenarios. Additionally, we further assess the robustness of each algorithm under varying conditions.

\subsection{Our Contributions}

Our study focuses on the problem of ANN search in hybrid query scenarios and provides a comprehensive review and evaluation of existing algorithms and systems. The main contributions are summarized in the following 4 aspects.

(1)\textbf{ Systematic Classification and Overview.}
We systematically classify 10 representative attribute-filtering algorithms from multiple dimensions, including index organization, filtering strategy, Boolean logic support, and index construction methods. Additionally, we survey range-filtering algorithms and provide an overview of widely used vector retrieval libraries (e.g., Faiss) and vector databases (e.g., Milvus, PASE, VBASE). These efforts provide a unified reference framework for future study.

(2)\textbf{ Evaluation of Attribute-Filtering Algorithms.}
We conduct a systematic evaluation of 13 attribute filtering algorithms on 6 real-world datasets. By analyzing performance under varying numbers of attributes, we reveal the strengths and weaknesses of each method. We further examine their behavior under different attribute distributions and selectivity conditions to assess robustness and adaptability in complex query scenarios. Evaluation metrics include index construction time, index size, peak memory usage, QPS, and search accuracy.

(3)\textbf{ Evaluation of Range-Filtering Algorithms.}
We benchmark 5 mainstream range-filtering algorithms on 3 large-scale datasets, using varying query range settings in the experiments. The experimental results demonstrate the performance of these algorithms in terms of index construction efficiency, storage overhead, and query performance. Additionally, we conduct an in-depth analysis of how index organization impacts algorithm performance, providing valuable insights for algorithm design.

(4)\textbf{ Recommendations and Challenges.}
Based on the experimental results, we provide algorithm selection recommendations for common application scenarios and highlight key challenges in the field of hybrid queries. These challenges include limited Boolean logic support, the lack of multi-attribute range-filtering capabilities, the high indexing costs of graph-based methods, and the sensitivity of algorithms to data distribution. Currently, few methods simultaneously support both attribute filtering and range filtering, pointing to potential directions for future research.

@@

However, to address the curse of dimensionality faced by NN search \cite{dimcurse}, existing studies focus on approximate solutions, known as ANN  search. We typically use $\text{Recall}@k = \frac{|R \cap \hat{R}|}{k}$ to evaluate the accuracy of ANN search algorithms, where $R$ denotes the true top-$k$ nearest neighbors of the query, and $\hat{R}$ denotes the approximate top-$k$ nearest neighbors returned by the ANN search algorithm.

As the complexity of real-world application requirements increases, NN search has evolved into hybrid NN search with attribute constraints. Depending on the nature of the attribute constraints, hybrid NN search can be divided into 2 categories: 1) Attribute Filtering Nearest Neighbor (\textit{AF-NN}) search. 2) Range Filtering Nearest Neighbor (\textit{RF-NN}) search. We provide their formal definitions below.

@@

\subsection{Index Organization in Hybrid ANN Search}

Current hybrid query methods mainly adopt graph-based ~\cite{nsw,kgraph,nsg,fanng,ngt} or Inverted File Index (IVF)-based \cite{PQ} index organization. We briefly introduce the fundamental principles of these 2 types in the following.

\vspace{1em}
\noindent\textbf{\underline{Graph.}}
Graph-based ANN search algorithms accelerate queries by building graph indexes on datasets. As illustrated in Figure~\ref{fig:graph}, each data point in the dataset corresponds to a point in the graph, and neighboring vertices (e.g., $a$ and $b$) are connected via edges based on their distance $\textit{dist}(a, b)$. In the graph index, each point maintains connections only to its nearest neighbors to preserve query efficiency. For instance, in Figure~\ref{fig:graph}, the gray point $c$ is connected to 5 neighbors $(a, b, d, e, f)$ and can traverse to them via corresponding edges.

Given a query point $q$, the graph-based ANN search typically follows a greedy search strategy to find the $k$ nearest neighbors. For illustration, consider the case where $k = 1$ (i.e., 1NN): The algorithm initiates the search from an entry point (the gray point $c$). If there exists a neighbor $n$ (the point $f$) of the current point such that $\textit{dist}(n, q) < \textit{dist}(c, q)$, the algorithm updates the current point to $n$. This process is iteratively repeated—always moving to the neighbor closest to $q$—until no neighbor is found that is closer than the current point. The final point reached (the blue point $j$) is returned as the approximate nearest neighbor of $q$.

@@

We provide an overview and analysis of 15 hybrid query methods, all drawn from recent studies, and collectively referred to as algorithms in the following. Specifically, Section~3.1 describes methods for attribute filtering, Section~3.2 discusses those for range filtering, and Section~3.3 introduces vector libraries and databases that support both attribute filtering and range filtering.

\vspace{-10pt}
\subsection{Attribute-Filtering Algorithms}

We summarize 6 attribute-filtering algorithms and provide an intuitive comparison in Table~\ref{tab:compair_1}.

\vspace{1em}
\noindent\textbf{\underline{NHQ \cite{NHQ}.}}  
Traditional attribute-filtering algorithms usually perform attribute constraints and ANN search separately. In contrast, NHQ is the first to implement simultaneous filtering, integrating both aspects into a unified framework. NHQ constructs an index based on a nearest neighbor graph and introduces a fusion distance that jointly captures vector similarity and attribute similarity. Leveraging this fusion distance, NHQ unifies vector similarity and attribute matching into a single comprehensive similarity measure and builds the graph accordingly. During the query process, NHQ efficiently prunes irrelevant edges via this composite index, enabling fast retrieval of results that satisfy both vector similarity and attribute constraints.

@@

Filtered-DiskANN also supports simultaneous filtering. Built upon the Vamana \cite{diskann} graph-based ANN index, it incorporates attribute information directly into the graph during index construction. This integration ensures that the index reflects both vector similarity and attribute constraints. Moreover, Filtered-DiskANN supports SSD-based storage, improving scalability.

Filtered-DiskANN proposes 2 index variants:  
1) \textbf{FilteredVamana}, which incrementally builds the graph index by inserting data points and dynamically adding edges, allowing adaptive expansion.  
2) \textbf{StitchedVamana}, which adopts a batch construction strategy—constructing separate Vamana subgraphs for each attribute, followed by merging and edge pruning.

@@

ACORN is built upon Hierarchical Navigable Small World (HNSW) \cite{hnsw}. It supports high-cardinality and unrestricted predicate sets, overcoming the limitations of methods restricted to small-scale equality filters. It constructs a denser hierarchical graph by expanding node neighborhoods and prunes edges at lower levels to control index size. During the query process, ACORN eliminates unnecessary distance computations by filtering out nodes that violate attribute constraints. This effectively maintains a nearest neighbor graph over valid nodes only. 
%During the query process, ACORN avoids unnecessary distance computations by filtering out nodes that do not meet attribute constraints, effectively simulating a nearest neighbor graph over valid nodes only.

ACORN includes 2 index variants:  
1) \textbf{ACORN-$\gamma$}, which expands neighbor lists during construction, increasing memory usage for higher performance.  
2) \textbf{ACORN-1}, which extends neighbor lists by considering second-hop neighbors during search, reducing index size with minor performance loss.

\vspace{1em}
\noindent\textbf{\underline{UNG \cite{UNG}.}} Graph-based methods (NHQ, Filtered-DiskANN, ACO-\\RN) do not guarantee result completeness, while CAPS lacks robustness when base and query vectors differ in attribute cardinality. UNG is proposed to overcome both limitations.

@@

UNG supports Boolean filtering with 2 modes:  
1) The query attribute set is a subset of the data attribute set.  
2) The query attribute set exactly matches the data attribute set.

@@

\subsection{Range-Filtering Algorithms}

We introduce 5 representative range-filtering algorithms and present a comparative overview in Table~\ref{tab:range_algo}.

\vspace{1em}
\noindent\textbf{\underline{SeRF \cite{serf}.}}  
Conventional RF-ANN search approaches typically follow one of two paradigms: 1) Conducting ANN search first followed by attribute-based filtering. 2) Filtering the dataset based on attribute ranges before performing ANN search. However, both approaches suffer from suboptimal performance. Building a separate neighbor graph (e.g., HNSW) for each attribute range could ensure efficient querying but incurs an $O(n^2)$ cost in constructing and storing $n$ graphs. Since many edges are shared across these graphs, SeRF introduces a validity-range aware design where each edge records the interval in which it is valid, indicating in which subgraphs the edge remains valid. This compresses $n$ graphs into a single unified graph, maintaining search effectiveness while significantly reducing memory and index construction overhead.
